\thetitle

### 2008-05-30 paper.tex
...

Paxson and Floyd observed~\cite{Paxson95} that the interplay between endpoint behavior and the network conditions is inherently \textit{closed-loop} in the sense that it is potentially affected by complex feedback. Traffic models typically attempt to preserve the closed-loop behavior of network traffic~\cite{Avallone04,Hernandez06:dissertation}. This presents a fundamental difficulty, however, in that it presumes that we know the intent of endpoints: what \textit{would} they have done under different conditions? While we can speculate about what an individual node might hypothetically do, we currently hardly understand the impact of the full-network traffic pattern upon performance metrics at all---even without trying to account for hypothetical reactions to alternate situations. Especially in the wireless setting, a fuller understanding of total network behavior must be reached before we can sensibly tackle the complexities of multi-level behavioral feedback. Accordingly, in this paper we attempt to provide a first-order approximation of complete network behavior by studying the response of performance metrics to \textit{open-loop} traffic models without multi-level feedback. It is important to realize that while this does not provide a final picture, we currently lack even a first-order understanding of the effect of different workloads on performance. This first-order understanding is an essential initial step.

There have been many studies of large wireless network deployments~\cite{Tang99,Balachandran02,Balazinska03,Kotz02,Henderson04,Schwab04,Chinchilla04,Jardosh05:ewind}. These analyses have described a wide variety of aspects of wireless network behavior, and provide much insight into the workings of real, deployed wireless networks. These studies present a broad analysis of general system features and trends of specific corporate wireless local-area networks (\caps{WLAN}s)~\cite{Tang99,Balazinska03}, university campus \caps{WLAN}s~\cite{Tang00,Kotz02,Chinchilla04,Schwab04,Henderson04,Tuduce05}, and temporary \caps{WLAN}s at conference venues~\cite{Balachandran02,Jardosh05:ewind}. They also provide a large body of raw data for subsequent analysis and modeling research. Our work provides the methodology for turning this rich foundation of field data into usable, realistic models of workload for a wide variety of networking situations.

The choice of mobility models for mobile wireless simulations can have a drastic impact on important performance metrics~\cite{Camp02,Yoon03:speed-decay,Yoon03:sound-models,Jardosh03,Zheng04}. Moreover, commonly used but simplistic mobility models, such as \caps{RWP}, exhibit characteristics, including density waves and speed decay, that are categorically dissimilar from any known real-world behavior~\cite{Royer01,Yoon03:speed-decay,Yoon03:sound-models}. In response to this evidence, more realistic mobility models have been proposed~\cite{Jardosh03,Tuduce05,Jardosh05:voronoi}. While much of this work focuses on making models that are simply more intuitively appealing~\cite{Jardosh03,Jardosh05:voronoi}, some work has begun to capitalize on this newly created wealth of wireless field data by deriving models from observed usage behavior, rather than intuition alone~\cite{Balazinska03,Tuduce05}.
%The analysis of Balazinska \textit{et al.}~\cite{Balazinska03} introduces the notions of the prevalence and persistence of an \caps{AP} with respect to mobile nodes.\footnote{These concepts are adapted from Paxson's work on Internet routing behavior: ``\textit{prevalence} meaning the overall likelihood that a particular route is encountered, and \textit{persistence}, the likelihood that a route remains unchanged over a long period of time''~\cite{Paxson96}.} Tuduce \textit{et al.}~\cite{Tuduce05} have based their mobility model on the persistence and prevalence characteristics of real trace data, which makes their work particularly relevant to our definition of sufficient realism. They do not, however, determine the impact of prevalence and persistence characteristics on performance metrics.

In this paper, instead of mobility, we examine an even more fundamental aspect of user behavior in wireless networks: the pattern of traffic generated by users and applications. This aspect of behavior is more fundamental because it applies to all types of wireless networks, not just mobile and ad~hoc networks. Moreover, the effect of traffic patterns applies not only to simulations, but also to experimental deployments, which have become the gold standard for wireless protocol evaluation.  Experimental deployments sidestep the problem of accurately modeling the lower layers of the network. Unless traffic and mobility are modeled realistically, however, the experimental results will still be unreliable.

>>>>

Paxson and Floyd observed~\cite{Paxson95} that the interplay between endpoint behavior and the network conditions is inherently \textit{closed-loop} in the sense that it is potentially affected by complex feedback. Traffic models typically attempt to preserve the closed-loop behavior of network traffic~\cite{Avallone04,Hernandez06:dissertation}. This presents a fundamental difficulty, however, in that it presumes that we know the intent of endpoints: what \textit{would} they have done under different conditions? While we can speculate about what an individual node might hypothetically do, we currently hardly understand the impact of the full-network traffic pattern upon performance metrics at all---even without trying to account for hypothetical reactions to alternate situations. A fuller understanding of total network behavior must be reached before we can sensibly tackle the complexities of multi-level behavioral feedback. Accordingly, in this paper we attempt to provide a first-order approximation of complete network behavior by studying the response of performance metrics to \textit{open-loop} traffic models without multi-level feedback. It is important to realize that while this does not provide a final picture, we currently lack even a first-order understanding of the effect of different workloads on performance. This first-order understanding is an essential initial step.

% There have been many studies of large wireless network deployments~\cite{Tang99,Balachandran02,Balazinska03,Kotz02,Henderson04,Schwab04,Chinchilla04,Jardosh05:ewind}. These analyses have described a wide variety of aspects of wireless network behavior, and provide much insight into the workings of real, deployed wireless networks. These studies present a broad analysis of general system features and trends of specific corporate wireless local-area networks (\caps{WLAN}s)~\cite{Tang99,Balazinska03}, university campus \caps{WLAN}s~\cite{Tang00,Kotz02,Chinchilla04,Schwab04,Henderson04,Tuduce05}, and temporary \caps{WLAN}s at conference venues~\cite{Balachandran02,Jardosh05:ewind}. They also provide a large body of raw data for subsequent analysis and modeling research. Our work provides the methodology for turning this rich foundation of field data into usable, realistic models of workload for a wide variety of networking situations.
% 
% The choice of mobility models for mobile wireless simulations can have a drastic impact on important performance metrics~\cite{Camp02,Yoon03:speed-decay,Yoon03:sound-models,Jardosh03,Zheng04}. Moreover, commonly used but simplistic mobility models, such as \caps{RWP}, exhibit characteristics, including density waves and speed decay, that are categorically dissimilar from any known real-world behavior~\cite{Royer01,Yoon03:speed-decay,Yoon03:sound-models}. In response to this evidence, more realistic mobility models have been proposed~\cite{Jardosh03,Tuduce05,Jardosh05:voronoi}. While much of this work focuses on making models that are simply more intuitively appealing~\cite{Jardosh03,Jardosh05:voronoi}, some work has begun to capitalize on this newly created wealth of wireless field data by deriving models from observed usage behavior, rather than intuition alone~\cite{Balazinska03,Tuduce05}.
%The analysis of Balazinska \textit{et al.}~\cite{Balazinska03} introduces the notions of the prevalence and persistence of an \caps{AP} with respect to mobile nodes.\footnote{These concepts are adapted from Paxson's work on Internet routing behavior: ``\textit{prevalence} meaning the overall likelihood that a particular route is encountered, and \textit{persistence}, the likelihood that a route remains unchanged over a long period of time''~\cite{Paxson96}.} Tuduce \textit{et al.}~\cite{Tuduce05} have based their mobility model on the persistence and prevalence characteristics of real trace data, which makes their work particularly relevant to our definition of sufficient realism. They do not, however, determine the impact of prevalence and persistence characteristics on performance metrics.

% In this paper, instead of mobility, we examine an even more fundamental aspect of user behavior in wireless networks: the pattern of traffic generated by users and applications. This aspect of behavior is more fundamental because it applies to all types of wireless networks, not just mobile and ad~hoc networks. Moreover, the effect of traffic patterns applies not only to simulations, but also to experimental deployments, which have become the gold standard for wireless protocol evaluation.  Experimental deployments sidestep the problem of accurately modeling the lower layers of the network. Unless traffic and mobility are modeled realistically, however, the experimental results will still be unreliable.
